In its opening chapters, this thesis started off with a very general overview of the myriad `pressures' and `biases' that have been proposed to influence language change. Chapter~\ref{ch:momentummodel} proposed a new pressure to be added to the pool of established ones, with Chapter~\ref{ch:questionnaire} adding to the existing empirical evidence that a momentum-based approach is indeed a viable candidate like other previously proposed pressures.

Having honed in on one particular pressure, this final chapter is intended to `zoom out' again to help situate the idea of momentum-based selection relative to other mechanisms that are involved in language change, as well as connect it to previous thinking in the literature. % situate it in its historical context
The goal is to draw the many pressures that were touched upon back together in a systematic way, to form a more wholesome, if more complex, picture of language change. As this is a complex issue, I will approach it from four different sides: Section~\ref{sec:competingmotivations} discusses the idea of \emph{competing pressures} and how it has been treated in the literature, as well as provide theoretical motivations for why momentum could (and should) be relevant for language change research. % TODO also Bickel stuff?

In order to get a more systematic grip on the `pressures' alluded to throughout this thesis, Section~\ref{sec:wrightfisher} will introduce the \emph{Wright-Fisher model} of biological evolution and investigate it using the Markov model toolkit that was already used in Section~\ref{markovmodel}. The main purpose of this is to allow us to study a toy model of the dynamics and interaction of two basic pressures that can be incorporated into it, namely \emph{mutation} and \emph{selection}. Having covered the basics of the Wright-Fisher model, Section~\ref{sec:asymmetricmutation} will provide an extension that implements a momentum-based selection bias.

Finally, Section~\ref{sec:conclusion} will provide an overall summary and conclusion of the thesis, as well as a look forward on future work.

% to situate the pressure in relation to other mechanisms that are involved in language change,

\section{Competing motivations in linguistic change}\index{competing motivations}
\label{sec:competingmotivations}
So far, the pressures were presented as directly set up against each other. But, partly in order to cope with the problem of \emph{actuation}, \index{actuation problem} scholars have already considered the interaction between them.

\citet[p.498]{Labov2001} states that his comprehensive studies of sound changes
\begin{quotation}
made it clear that linguistic change in progress is heavily constrained by the physical environment in which it takes place, and by structural factors that limit the course of change. At the same time, it is argued that the forces that move and motivate change, and are responsible for incrementation and transmission across generations, are largely social in nature.
\end{quotation}

\subsection{Relating the pressures}

Referring back to the two \emph{rates of change} discussed in Section~\ref{sec:discussion}, it becomes obvious that two different subfields have studied the same phenomenon on two very different levels:\index{rates of change} work in diachronic typology~\citep[such as][]{Bickel2015} is interested in the actuation probabilities on a macro-level, whereas traditional historical linguistic work hones in on particular changes.

% does this mean that language always has a reason to move in many different directions? why is it ever standing still then?

\citet{Sundgren2009}: The varying influence of social and linguistic factors on language stability and change

\citet{Wichmann2015}: Diachronic stability and typology (\citet{Nichols2003} chapter from earlier version of same handbook?)

historical contingency~\citep[p.503]{Labov2001} % Gould?

Ok so if there's already so much work on different pressures, why should we need momentum-based selection? Apart from the empirical reasons (which still need plenty of filling in), there are also theoretical reasons why we should \emph{a priori} prefer a mechanism like momentum-based selection over functional ones.

\subsection{Language change, arbitrariness, replicator-neutrality}

% arbitrariness
% directed variation + directional selection
\citet{Sapir1921}'s `drift': languages do not just diverge when there is social pressure to discriminate, but also have an internal `drive' to change (in place) by themselves (p.160).

\citep[ch.7-12]{Labov2010} on forks in the road etc.

challenging the notion that ``a reliance on chance deprives an account of any real explanatory value" \citep[p.44]{Joseph2013}.

\subsection{Neutral selection of non-random variation}

Evolang: not \emph{easier to learn} but \emph{more likely to be invented}

% cf. "the prevalence of a particular species in a habitat does not necessarily imply that is it any better adapted to that habitat than its competitors"~\citep[p.2]{Blythe2012copying}

% biology: \citep{Gerlee2015}

\subsection{Requirements of a theory of language change}

% \citep{Weinreich1968}

What do you want from a `theory of language change'? It shouldn't be able to accurately predict which changes are gonna happen where, but neither should it be completely not predicting anything. What it should be able to predict is \emph{the degree of its own unpredictability}, i.e. its own limitations. It should at the same time predict the distribution or likelihood of different types of changes overall (`actuation probabilities' if you so wish)\index{actuation!actuation probability} while \emph{explicitly underspecifying} its own predictions about whether a specific change is about to happen in a language or not.
% a model that predicts \emph{spontaneous actuations} (`noise-activated transitions' in the terms of \citealt{Mitchener2016})\index{actuation}

\input{conclusion/asymmetricvariation}

\section{Summary, conclusion \& outlook}

Evolutionary approach and explicit treatment of the generation of variants and their selection as seperate processes could inform the systematic cross-linguistic study of language changes such as \citet{Bickel2015}.% "very specific effects can be studied *without* having to rule out all the intervening effects that you find in such an exceedingly complex phenomenon such as language"
